---
title: 'IBM HR Analysis
author: "Sahas Duggirala"
output:
  pdf_document:
    toc: yes
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show

```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, warning=FALSE, message=FALSE, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(fivethirtyeight)
library(here)
library(skimr)
library(janitor)
library(vroom)
library(tidyquant)
library(rvest)    # scrape websites
library(purrr)  
library(lubridate) #to handle dates
```


IBM HR Analytics

First let us load the data

```{r}

hr_dataset <- read_csv(here::here("data", "datasets_1067_1925_WA_Fn-UseC_-HR-Employee-Attrition.csv"))
glimpse(hr_dataset)

```

I am going to clean the data set, as variable names are in capital letters, some variables are not really necessary, and some variables, e.g., `education` are given as a number rather than a more useful description


```{r}

hr_cleaned <- hr_dataset %>% 
  clean_names() %>% 
  mutate(
    education = case_when(
      education == 1 ~ "Below College",
      education == 2 ~ "College",
      education == 3 ~ "Bachelor",
      education == 4 ~ "Master",
      education == 5 ~ "Doctor"
    ),
    environment_satisfaction = case_when(
      environment_satisfaction == 1 ~ "Low",
      environment_satisfaction == 2 ~ "Medium",
      environment_satisfaction == 3 ~ "High",
      environment_satisfaction == 4 ~ "Very High"
    ),
    job_satisfaction = case_when(
      job_satisfaction == 1 ~ "Low",
      job_satisfaction == 2 ~ "Medium",
      job_satisfaction == 3 ~ "High",
      job_satisfaction == 4 ~ "Very High"
    ),
    performance_rating = case_when(
      performance_rating == 1 ~ "Low",
      performance_rating == 2 ~ "Good",
      performance_rating == 3 ~ "Excellent",
      performance_rating == 4 ~ "Outstanding"
    ),
    work_life_balance = case_when(
      work_life_balance == 1 ~ "Bad",
      work_life_balance == 2 ~ "Good",
      work_life_balance == 3 ~ "Better",
      work_life_balance == 4 ~ "Best"
    )
  ) %>% 
  select(age, attrition, daily_rate, department,
         distance_from_home, education,
         gender, job_role,environment_satisfaction,
         job_satisfaction, marital_status,
         monthly_income, num_companies_worked, percent_salary_hike,
         performance_rating, total_working_years,
         work_life_balance, years_at_company,
         years_since_last_promotion)

```

Below are a non-exhaustive list of questions we aim to tackle in the next few chunks of code.

1. How often do people leave the company (`attrition`)
1. How are `age`, `years_at_company`, `monthly_income` and `years_since_last_promotion` distributed? can you roughly guess which of these variables is closer to Normal just by looking at summary statistics? 
1. How are `job_satisfaction` and `work_life_balance` distributed? Don't just report counts, but express categories as % of total
1. Is there any relationship between monthly income and education? Monthly income and gender?
1. Plot a boxplot of income vs job role. Make sure the highest-paid job roles appear first
1. Calculate and plot a bar chart of the mean (or median?) income by education level.
1. Plot the distribution of income by education level. Use a facet_wrap and a theme from `ggthemes`
1. Plot income vs age, faceted by `job_role`

```{r}
# 1. We convert count data into percent to better understand the share of instances where people leave the company 

attrition_no <- hr_cleaned %>% 
 count(attrition) %>% 
  filter(attrition == "No") %>% 
  mutate(percent = (n/(1233+237)*100))  

  attrition_yes <- hr_cleaned %>% 
  count(attrition) %>% 
  filter(attrition == "Yes") %>% 
  mutate(percent = (n/(1233+237)*100)) 
  
  bind_rows(attrition_yes, attrition_no)


```

```{r}
# 2. We plot the distribution of various demographic parameters of the workforce

hr_cleaned %>% 
  summarise(mean(age), mean(years_at_company), mean(monthly_income), mean(years_since_last_promotion))

 ggplot(hr_cleaned, aes(x =age)) +geom_density(kernel="gaussian") + theme_economist() 
 ggplot(hr_cleaned, aes(x =years_at_company)) +geom_density(kernel="gaussian") +theme_economist() 
 ggplot(hr_cleaned, aes(x =monthly_income)) +geom_density(kernel="gaussian") +theme_economist() 
 ggplot(hr_cleaned, aes(x =years_since_last_promotion)) +geom_density(kernel="gaussian") +theme_economist() 

```


```{r}
# 3. Job Satisfaction and Work/Life balance distributions

JS <- hr_cleaned %>% 
  select(job_satisfaction)


JS_prop <- JS %>% 
  count(job_satisfaction) %>%    
  mutate(prop = prop.table(n))

ggplot(JS_prop, aes(x =job_satisfaction, y = prop)) + geom_col()

WLB <- hr_cleaned %>% 
  select(work_life_balance)


WLB_prop <- WLB %>% 
  count(work_life_balance) %>%    
  mutate(prop = prop.table(n))

ggplot(WLB_prop, aes(x =work_life_balance, y = prop)) + geom_col()



```


```{r}
 # 4. Exploring the relationship between income vs. education and income vs. gender

ggplot(hr_cleaned, aes(x=education, y=monthly_income)) +geom_boxplot()
ggplot(hr_cleaned, aes(x=gender, y=monthly_income)) +geom_boxplot()


```

```{r}
# 5. Mapping he relationship between job role and monthly income
ggplot(hr_cleaned, aes(x=monthly_income, y=reorder(job_role, monthly_income))) + geom_boxplot() 

```

```{r}
# 6. Mean and median income as per education level distributions 

Hr2 <- hr_cleaned %>% 
  group_by(education) %>% 
  mutate(mean_inc = mean(monthly_income)) %>% 
  mutate(median_inc = median(monthly_income))
  

  ggplot(Hr2, aes(x = mean_inc)) +geom_bar()
  ggplot(Hr2, aes(x = median_inc)) +geom_bar()

```


```{r}
 # 7. We plot the distribution of income by educational level
ggplot(hr_cleaned, aes(x = monthly_income)) + geom_dotplot() + theme_economist() +facet_wrap(~education)
```

```{r}

# 8. We plot the relationship of age vs income and we facet by job position 
 ggplot(hr_cleaned, aes(x = age, y = monthly_income)) + 
  geom_point() + 
  facet_wrap(~job_role) +
  theme_economist() 

```
