---
title: 'Health Data Analysis'
author: "Sahas Duggirala"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
  pdf_document:
    toc: yes
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
```




```{r, get_data, cache=TRUE}

# load gapminder HIV data
hiv <- read_csv(here::here("data","adults_with_hiv_percent_age_15_49.csv"))
life_expectancy <- read_csv(here::here("data","life_expectancy_years.csv"))

# get World bank data using wbstats
indicators <- c("SP.DYN.TFRT.IN","SE.PRM.NENR", "SH.DYN.MORT", "NY.GDP.PCAP.KD")


library(wbstats)

worldbank_data <- wb_data(country="countries_only", #countries only- no aggregates like Latin America, Europe, etc.
                          indicator = indicators, 
                          start_date = 1960, 
                          end_date = 2016)

# get a dataframe of information regarding countries, indicators, sources, regions, indicator topics, lending types, income levels,  from the World Bank API 
countries <-  wbstats::wb_cachelist$countries

view(countries)
view(hiv)
view(life_expectancy)
view(worldbank_data)

```

You have to join the 3 dataframes (life_expectancy, worldbank_data, and HIV) into one. You may need to tidy your data first and then perform [join operations](http://r4ds.had.co.nz/relational-data.html). Think about what type makes the most sense **and explain why you chose it**.

``` {r, tidydata}

tidy_life_expectancy <- life_expectancy %>% 
  pivot_longer(names_to="year", values_to="life_expect", cols= 2:302) %>% 
  transform(year = as.numeric(year)) %>% 
  drop_na(life_expect)


tidy_hiv_data <- hiv %>% 
  pivot_longer(names_to="year", values_to ="hiv_proportion", cols= 2:34) %>% 
  transform(year = as.numeric(year)) %>%
  drop_na(hiv_proportion)  


tidy_world_bank <- worldbank_data %>% 
  rename("GDP_per_capita"=NY.GDP.PCAP.KD, 
         "School_enrollment"= SE.PRM.NENR, 
         "Mortality_rate"=SH.DYN.MORT,
         "Fertility_rate" =SP.DYN.TFRT.IN, 
         "year"=date)

# after cleaning the data, we will join the relevant pieces of data. 


joined_data_1 <- full_join(tidy_hiv_data, tidy_life_expectancy, by = c('country','year'))
joined_data_2 <- full_join(joined_data_1, tidy_world_bank, by = c('country','year'))

joined_data_2


```
``` {r, joindata}

country_data <- left_join(joined_data_2, countries, by = "country")

country_data
tidy_country_data <- country_data %>% 
drop_na()
  
tidy_country_data


```


1. What is the relationship between HIV prevalence and life expectancy? 

``` {r, HIV_Life}

ggplot(tidy_country_data, aes(y=life_expect, x=hiv_proportion, na.rm=TRUE))+ 
  geom_point(alpha=0.1)+
  geom_smooth(size=0.5)+
  facet_wrap(~region,scales = "free")+ 
  theme_economist()+
  theme(axis.title = element_text())+
  labs(y="Life Expectancy (Years)", x= "HIV Proportion (%)", title="HIV Rate and Life Expectancy")

```


1. What is the relationship between fertility rate and GDP per capita? 

```{r, Fertility_GDP}

ggplot(tidy_country_data, aes(y=GDP_per_capita, x=Fertility_rate, na.rm=TRUE))+ 
  geom_point(alpha=0.1, na.rm=TRUE)+
  geom_smooth(na.rm=TRUE)+
  facet_wrap(~region,scales = "free")+ 
  theme_economist()+
  theme(axis.title = element_text())+
  labs(y="GDP Per Capita ($)", x= "Fertility Rate (%)", title="GDP Per Capita and Fertility Rate")


```



1. Which regions have the most observations with missing HIV data? 

```{r, missing}
missing_hiv_data <-country_data %>% 
  select("hiv_proportion","region") %>% 
  drop_na(region) %>% 
  filter(is.na(hiv_proportion)) %>% 
  group_by(region) %>% 
  summarise(count =n()) %>% 
  arrange(desc(count))

missing_hiv_data

library(kableExtra)


kbl(missing_hiv_data,col.names=c("Region","Number of NA Values in HIV Data")) %>%
  kable_styling()


ggplot(missing_hiv_data, aes(x= count, y=reorder(region, count)))+
  geom_col()+
  labs(x="Number of Missing Entries", y="region", title="Number of missing entries for HIV data")+
  theme_economist()+
  theme(axis.title=element_text())

```


1(a). How has mortality rate for under 5 changed by region? 
1(b). Is there a relationship between primary school enrollment and fertility rate?

```{r,improvement_data}

# mortality rate by region 

mortality_plot <- tidy_country_data %>%
  group_by(region,year) %>% 
  summarize(mean_mortality=mean(Mortality_rate))
mortality_plot


mortality_region <- ggplot(mortality_plot, aes(x=year, y=mean_mortality, group=region)) + 
  geom_line() + 
  facet_wrap(~region,scales='free')+ 
  theme_economist()+
  theme(axis.text.x=element_text())+
  labs(title = "Change in mortality rate over time")+
  xlab("Years") +
  ylab("Mortality rate")

mortality_region

```

