---
title: 'Session 6: Homework 3'
author: "Group B 21"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(tidyquant)
library(infer)
library(openintro)
library(tidyquant)
```


# Omega Group plc- Pay Discrimination


At the last board meeting of Omega Group Plc., the headquarters of a large multinational company, the issue was raised that women were being discriminated in the company, in the sense that the salaries were not the same for male and female executives. A quick analysis of a sample of 50 employees (of which 24 men and 26 women) revealed that the average salary for men was about 8,700 higher than for women. This seemed like a considerable difference, so it was decided that a further analysis of the company salaries was warranted. 


## Loading the data


```{r load_omega_data}
omega <- read_csv(here::here("data", "omega.csv"))
glimpse(omega) # examine the data frame
```

## Relationship Salary - Gender ?

Is there a significant difference between the salaries of the male and female executives?

```{r, confint_single_valiables}
# Summary Statistics of salary by gender
mosaic::favstats (salary ~ gender, data=omega)

# Dataframe with two rows (male-female) and having as columns gender, mean, SD, sample size, 
# the t-critical value, the standard error, the margin of error, 
# and the low/high endpoints of a 95% condifence interval
sum_stat <- omega %>% group_by(gender) %>%
            summarise(mean=mean(salary),
                      sd=sd(salary),
                      samp_size=n(),
                      t_crit= qt(.975,samp_size-1),
                      se=sd/sqrt(samp_size),
                      MoE=t_crit*se,
                      lower_ci_95 = mean-se,
                      upper_ci_95 = mean+se)
sum_stat
                  
```
>
Explanation: The summarized stats show that mean salary is lower for women compared to men salaries. The confidence intervals do not overlap and the difference is significant with 95% confidence..

Let's run a hypothesis test to be absolutely certain......

```{r, hypothesis_testing}
# hypothesis testing using t.test() 
t.test((omega %>% filter(gender=="male"))$salary,(omega%>%filter(gender=="female"))$salary)

# hypothesis testing using infer package
set.seed(1234)
simu <- omega %>% specify(salary~gender) %>% hypothesize(null="independence")%>%generate(reps=1000,type="permute")

sim_delta_mean <-simu%>% calculate(stat="diff in means", order=c("male","female"))

real_delta_mean <- omega %>% specify(salary~gender)%>% calculate(stat="diff in means", order=c("male","female"))

sim_delta_mean %>% visualise (obs_stat = real_delta_mean, direction="both")+labs(title="Wage gap by gender",subtitle="Observed difference in red")  

```

> 
Explanation: The observed gap is within the interval so we can reject the null hypothesis. Men and women do not make the same amount of money. This could be attributable to a variety of factors such as: Are there more men in sectors that pay more (eg: tech, finance) and this is skewing the data? We could go deeper to analyze the proportion of men and who choose STEM degrees and what the average pay for a STEM graduate is. Furthermore, specitic to Omega, are there more men in senior positions? Department-level data should be used versus the organization as a whole. 

## Relationship Experience - Gender?

Is there a correlation between Experience on the job and gender?

```{r, experience_stats}
# Summary Statistics of salary by gender
favstats (experience ~ gender, data=omega)

```


>
Explanation: The difference in experience between male and female executives may account for the difference in wages but requires testing to confirm the correlation.


## Relationship Salary - Experience ?

What's the relationship between salary and experience?


```{r, salary_exp_scatter}
ggplot(omega, aes(experience,salary)) + geom_point() +geom_smooth(method=lm) +
  labs(x="Exp (years)", y = "Salary")

```


Let's check correlations between the data


```{r, ggpairs}
omega %>% 
  select(gender, experience, salary) %>% #order variables they will appear in ggpairs()
  ggpairs(aes(colour=gender, alpha = 0.3))+
  theme_bw()
```

Explanation: The experience has a stronger correlation with salary for female group.

